A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-01-30 analysis/paper/paper.Rmd
added datasets table, update to methods section

---
title: "A comparison of single-cell trajectory inference methods: towards more accurate and robust tools"
author:
  - Wouter Saelens* ¹ ²
  - Robrecht Cannoodt* ¹ ² ³
  - Helena Todorov¹ ² ⁴ ⁵
  - Yvan Saeys¹ ²
date: "`r format(Sys.time(), '%d %B, %Y')`"
bibliography: references.bib
csl: nature-biotechnology.csl
abstract: |
  Using single-cell transcriptomics data, it is now possible to computationally order cells along trajectories, allowing the unbiased transcriptome-wide study of cellular dynamic processes. Since 2014, already dozens of trajectory inference methods have been developed, each with its own set of methodological characteristics. This makes a direct comparison between trajectory inference methods often difficult, and as a result a comprehensive assessment of the performance and robustness of each method is still lacking. We developed a framework of data structures for trajectory inference methods in order to make their results comparable to each other and to a gold standard, when available. In this work, we compare the results from a total of 22 trajectory inference methods, on a large collection of real datasets and realistic synthetic datasets. We employ a novel parameter optimisation procedure, and compare methods using several metrics, including accuracy of the inferred ordering, correctness of the network wiring, code quality and user friendliness. We conclude with practical guidelines for method users, and also believe our evaluation framework can be used to spearhead the development for new methods. Our evaluation pipeline is fully reproducible, can be easily extended, and is available at www.github.com/dynverse.
keywords: |
  trajectory inference; pseudotemporal ordering; single-cell transcriptomics
highlights: |
  These are the highlights. 
---
\* Equal contribution  
¹ Data mining and modelling for biomedicine, VIB Center for Inflammation Research, Ghent, Belgium.  
² Department of Applied Mathematics, Computer Science and Statistics, Ghent University, Ghent, Belgium.  
³ Center for Medical Genetics, Ghent University Hospital, Ghent, Belgium.  
⁴ CIRI, Centre International de Recherche en Infectiologie, Inserm, Lyon, France.  
⁵ Université Claude Bernard Lyon 1, Lyon, France.  
<!----
Tips for working in this document:
* Go to tools -> perferences and remove Use smart quotes
* Add extra enters after every paragraph (this is markdown!)

@@

```{r setup, include=FALSE}
source("setup.R")
```

@@

@@

We created a transparent scoring scheme to check the quality of each method (`r ref("table", "methods")`) and weighed each quality check by the number of times it was found in a list of

@@

We gathered

@@

# Acknowledgements

# Supplementary

@@

```{r}
# what commit is this file at? You may need to change the path value
# if your Rmd is not in analysis/paper/
git2r::repository(".")
```

# References

[a]Outdated

>>>>

---
title: "A comparison of single-cell trajectory inference methods: towards more accurate and robust tools"
bibliography: references.bib
csl: nature-biotechnology.csl
---

*`r format(Sys.time(), '%d %B, %Y')`*
<!----
Tips for working in this document:
* Go to tools -> perferences and remove Use smart quotes
* Add extra enters after every paragraph (this is markdown!)

@@

```{r setup, include=FALSE}
source("setup.R")
```
Wouter Saelens* ¹ ² , Robrecht Cannoodt* ¹ ² ³, Helena Todorov¹ ² ⁴ ⁵, Yvan Saeys¹ ²

\* Equal contribution  
¹ Data mining and Modelling for Biomedicine, VIB Center for Inflammation Research, Ghent, Belgium.  
² Department of Applied Mathematics, Computer Science and Statistics, Ghent University, Ghent, Belgium.  
³ Center for Medical Genetics, Ghent University Hospital, Ghent, Belgium.  
⁴ CIRI, Centre International de Recherche en Infectiologie, Inserm, Lyon, France.  
⁵ Université Claude Bernard Lyon 1, Lyon, France.

# Abstract

Using single-cell transcriptomics data, it is now possible to computationally order cells along trajectories, allowing the unbiased transcriptome-wide study of cellular dynamic processes. Since 2014, already dozens of trajectory inference methods have been developed, each with its own set of methodological characteristics. This makes a direct comparison between trajectory inference methods often difficult, and as a result a comprehensive assessment of the performance and robustness of each method is still lacking. We developed a framework of data structures for trajectory inference methods in order to make their results comparable to each other and to a gold standard, when available. In this work, we compare the results from a total of 22 trajectory inference methods, on a large collection of real datasets and realistic synthetic datasets. We employ a novel parameter optimisation procedure, and We compare methods using several metrics, including accuracy of the inferred ordering, correctness of the network wiring, code quality and user friendliness. We conclude with practical guidelines for method users, and also believe our evaluation framework can be used to spearhead the development for new methods. Our evaluation pipeline is fully reproducible, can be easily extended, and is available at https://www.github.com/dynverse.

@@

```{r}
add_stable(
read_rds(figure_file("datasets.rds","dataset_characterisation")), 
"datasets", 
"Real datasets used in this study"
)
```

@@

We created a transparent scoring scheme to check the quality of each method (`r ref("table", "methods")`) and weighed each quality check by the number of times it was found in a list of references.

We gathered a list of real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`) .

@@

@@

# Acknowledgements
# References

<div id="refs"></div>

# Supplementary Material

@@

```{r}
# what commit is this file at? You may need to change the path value
# if your Rmd is not in analysis/paper/
git2r::repository(".")
```
[a]Outdated
